Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2020-12-03 bbbq_article.tex
Determine if test is run on a CI server, instead of just Travis

The human immune system is unexpectedly attentive towards
human transmembrane helices (TMHs), as TMH-derived peptide fragments
bind to histocompatibility complex (MHC) class I more often then expected
by chance only, which can be mostly explained by the low hydrophobicity
of those residues.
The physiological reason is yet unclear,
yet this finding hints that there has been selection upon detecting TMH-derived
peptides in pathogens.
If this signature of selection is present in MHC-II is unknown.
This study shows that MHC-I [also has/does not have] more
epitopes derived from a TMH for a pathogen proteome, when compared with
a host proteome.
Additionally, MHC-II binds to peptides derived from TMHs 
[less/equally/more] often than expected by chance.
Lastly, we show the TMHs are [more/less/equally] 
conserved.
Our findings suggest that the immune system is [less/neutral/more]
vigilant to TMHs than expected by chance and this has 
left [a clear/a weak/no]
signal in the evolutionary history of the pathogen.
The impact of this is [revolutionary/awesome].

@@

To detect the evolutionary conservation of TMHs,
we first selected all unique protein sequences per
gene from all 45k SARS-CoV-2 proteomes,
resulting in around 500 unique protein sequences per gene.
From the unique protein sequences, we made an alignment per gene,
using the msa (\cite{bodenhofer2015msa}) R package, 
set up to use ClustalOmega (\cite{thompson1994clustal})
with the BLOSUM80 amino acid transition matrix (\cite{henikoff1992amino}).
Per alignment, we used msa to determine the consensus sequence
and the conservation scores.
Per consensus sequences, the topology was determined using pureseqtmr.

\paragraph{Statistics}

We compare the conservation values for TMH versus non-TMH locations,
for all membrane proteins combined,
using a Kolmogorov-Smirnov test, for a significance level of 5 percent.

>>>>

The human immune system is unexpectedly attentive towards
human transmembrane helices (TMHs), as TMH-derived peptide fragments
bind to histocompatibility complex (MHC) class I more often then expected
by chance only, which can be mostly explained by the low hydrophobicity
of those residues.
The physiological reason is yet unclear,
yet this finding hints that there has been selection upon detecting TMH-derived
peptides in pathogens.
If this signature of selection is present in MHC-II is unknown.
This study shows that MHC-I [also has/does not have] more
epitopes derived from a TMH for a pathogen proteome, when compared with
a host proteome.
Additionally, MHC-II binds to peptides derived from TMHs 
[less/equally/more] often than expected by chance.
Lastly, we show the TMHs are [more/less/equally] 
conserved by determining if SNPs are found in TMH more/less/equal
than expected by chance.
Our findings suggest that the immune system is [less/neutral/more]
vigilant to TMHs than expected by chance and this has 
left [a clear/a weak/no]
signal in the evolutionary history of the pathogen.
The impact of this is [revolutionary/awesome].

@@

To detect the evolutionary conservation of TMHs, 
we use the follow workflow (also shown
in figure \ref{fig:snp_workflow}) to be able to count the
number of SNPs that occur in TMHs, as well as the number of SNPs
that occur in TMHs by chance.

To do so, we use the 
NCBI (\url{https://www.ncbi.nlm.nih.gov/}) databases:
'gene' to find the gene names of membane proteins, 
'dbSNP' for SNPs associated with those genes
and 'protein' to obtain the sequence of proteins that SNPs act upon.
The 'dbSNP' contains 2 billion (human-only) SNPs, although
it is estimated that approx 10 percent is a false positive 
\richel{see references on dbSNP Wikipedia page}.

The \verb;rentrez; R package provides for doing so from script.

The first query is a call to the 'gene' database for the 
term 'membrane protein' (in all fields) for the organism Homo sapiens.
This results in \richel{1126} gene IDs.
The next query is a call to the 'gene' database 
to obtain the gene names from the gene IDs.
Per gene name, the 'dbSNP' NCBI database is queried for 
SNPs associated with the gene name. The number of SNPs
was limited to the first \richel{100} SNPs per gene,
resulting in \richel{112600} SNPs.
Per selected SNP, all variation (i.e. descriptions of the changes on DNA
and/or RNA and/or protein level) are obtained from the 'dbSNP' database
as well. Of these variations, only the variations that result in a single
amino acid substitution are kept, resulting in \richel{???} protein changes.
Per protein substitution (i.e. a protein name and an index of the changed residue), 
the 'protein' NCBI database was queried for the
protein sequence.
Of each protein sequence, the protein topology is determined 
using PureseqTM.
From the topology and the known location of the amino acid substitution, 
it is tallied if the change occurred within a TMH or not.
Also, the percentage of the topology that is TMH is measured.
Not all membrane proteins have a transmembrane helix, resulting
in \richel{???} proteins with at least one TMH.

\begin{figure}[!htbp]
  \includegraphics[height=0.9\textheight]{snp_workflow.png}
  \caption{
    Workflow to determine the evolutionary conservation of TMHs
    using human SNPs.
  }
  \label{fig:snp_workflow}
\end{figure}

% \paragraph{Introduction}
% 
% To detect the evolutionary conservation of TMHs,
% we first selected all unique protein sequences per
% gene from all 45k SARS-CoV-2 proteomes,
% resulting in around 500 unique protein sequences per gene.
% From the unique protein sequences, we made an alignment per gene,
% using the msa (\cite{bodenhofer2015msa}) R package, 
% set up to use ClustalOmega (\cite{thompson1994clustal})
% with the BLOSUM80 amino acid transition matrix (\cite{henikoff1992amino}).
% Per alignment, we used msa to determine the consensus sequence
% and the conservation scores.
% Per consensus sequences, the topology was determined using pureseqtmr.
% 
% \paragraph{Statistics}
% 
% We compare the conservation values for TMH versus non-TMH locations,
% for all membrane proteins combined,
% using a Kolmogorov-Smirnov test, for a significance level of 5 percent.
